Efficient Thresholded Correlation using Truncated Singular Value Decomposition

### 2015-11-28 threshold/draft.Rnw
fixed a few typos, still working out examples

The number of correlation coefficients grows quadratically with the number of
vectors. Simply computing all pairs of correlation coefficients and then
filtering out coefficients below a given threshold may not be computationally
feasable for high-dimensional data in modern genomics and other applications.
Considerable attention has been devoted to \emph{pruning
methods} that cheaply prune away all but the most highly-correlated pairs of
vectors.  For example the TAPER and related algorithms of Xiong et al. develop
pruning rules based on the frequency of mutually co-ocurring observations (rows)
among vector pairs; see \cite{prune1, prune2} and the references therein.
Related methods approximate a set containing the most highly-correlated
pairs using hashing algorithms\cite{prune3}. A number of other methods arrive
at approximate sets of the most highly-correlated pairs using
dimensionality-reduction techniques like the discrete Fourier
transform\cite{timeseries}. The method of Wu, et al.\cite{svd-similarity},
conceptually very similar to our approach, uses the singular value
decomposition (SVD) to find all pairs of close vectors with respect to a
distance metric (instead of correlation).  Our method finds all of the most
highly-correlated vector pairs with respect to a given threshold by pruning
along an intuitive path of decreasing variance defined by the data and computed
by a truncated SVD.

@@

We evaluted the algorithm using three publically-available gene expression
datasets.  The tests were performed on a desktop PC with a single AMD A10-7850K
3.7$\,$GHz Athlon quad-core CPU and 16$\,$GB 1333$\,$MHz unbuffered RAM running
the Ubuntu 12.04.5 GNU/Linux OS. Tests were performed with 64-bit R version
3.1.2 using double-precision arithmetic and the OpenBLAS library version
0.1alpha2.2-3 (based on Goto's BLAS version 1.13) with {\tt
OMP\_NUM\_THREADS=4}.  All examples use the reference R implementation
function, {\tt tcor()}, which can be installed for the R language from the
development GitHub repository using the {\tt devtools} package with:

@@

Anyone can quickly run the expample to experiment with the algorithm and
compare its output with brute force results. The example uses the {\tt
BicatYeast} data from the {\tt biclust} R package and related {\tt BicAT}
software\cite{biclust, bicat}. The data consist of 70 experiments invovling the
Saccharomyces Cerevisiae yeast measuring gene expression data across 419
Affymetrix probes, represented as a $70\times 419$ real-valued matrix.  The
example finds all pairs of columns (Affymetrix probes) that meet or exceed a
Pearson correlation value of 0.95.  We use a projected dimension $p=10$ in {\tt tcor()}.
\begin{small}
<<eval=TRUE,echo=TRUE>>=
library(irlba)
source("tcor.R")  # XXX Replace later with tcor pacakge
library(biclust)  # for the BicatYeast data
data(BicatYeast)  # in gene by sample orientation
A = t(BicatYeast) # to correlate genes to genes
threshold = 0.95

@@

<<echo=FALSE, results=tex>>=
# --------------------------------------------------------------------------------------------------
# The first time you run this, you need to download the data as outlined in the
# following commented-out section. This code creates a file named
# 'rpkm.RData' used subsequently.
#
# To run any of the remaining (large) examples, first set the system environment
# variable RUN_EXAMPLES=TRUE (otherwise they won't run and representative results
# from Bryan's test machine will be printed).
#
#library(doMC)
# Download and extract this file with, for example:
# wget  http://gdac.broadinstitute.org/runs/stddata__2015_08_21/data/BRCA/20150821/gdac.broadinstitute.org_BRCA.Merge_rnaseq__illuminahiseq_rnaseq__unc_edu__Level_3__gene_expression__data.Level_3.2015082100.0.0.tar.gz
# tar zxf gdac.broadinstitute.org_BRCA.Merge_rnaseq__illuminahiseq_rnaseq__unc_edu__Level_3__gene_expression__data.Level_3.2015082100.0.0.tar.gz
# cd gdac.broadinstitute.org_BRCA.Merge_rnaseq__illuminahiseq_rnaseq__unc_edu__Level_3__gene_expression__data.Level_3.2015082100.0.0
# R --slave < read.R   # (run this file)
#f = "BRCA.rnaseq__illuminahiseq_rnaseq__unc_edu__Level_3__gene_expression__data.data.txt"
#N = 20534  # total number of lines
#h = 2      # header lines
# Read the first header line with sampe IDs:
#id = as.vector(read.table(f, sep="\t", stringsAsFactors=FALSE, header=FALSE, nrows=1))
# Read the 2nd header line with sample type (raw, RPKM, etc.)
#header = as.vector(read.table(f, sep="\t", stringsAsFactors=FALSE, skip=1, header=FALSE, nrows=1))
# Read the rest of the file in parallel. I prefer doMC/foreach over mclapply
# for this because foreach can combine results incrementally.
#cores = 4
#registerDoMC(cores)
#block = floor((N - h)/cores)
#t1 = proc.time()
#ans = foreach(j=1:cores, .combine=rbind) %dopar%
#{
#    skip = block * (j - 1) + h
#    nrows = ifelse(j == cores, -1, block)
#    read.table(f, sep="\t", stringsAsFactors=FALSE, skip=skip, header=FALSE, nrows=nrows)
#}
#names(ans) = header
#print(proc.time() - t1)
# (All timings shown for a quad-core Athlon home PC reading from an older
# OCZ REVODRIVE3 X2 SSD drive (ca 2014).)
#   user  system elapsed
#494.549   4.134 131.232
# Compare with sequential approach:
#t1 = proc.time()
#x = read.table(f, sep="\t", stringsAsFactors=FALSE, skip=1, header=TRUE)
#print(proc.time() - t1)
#   user  system elapsed
#517.430   1.318 519.553
# Create a samples x gene matrix of RPKM values
#genes = ans[, 1] # gene labels
#idx   = grep("RPKM", names(ans))    # Select columns with RPKM values
#rpkm     = t(as.matrix(ans[, idx]))
# Label the coordinates by sample ID and gene:
#rownames(rpkm) = id[idx]
#colnames(rpkm) = genes
#save(rpkm, file="rpkm.RData")
# --------------------------------------------------------------------------------------------------
# Default values:
t1=40.34
t2=35.60
m1=39.10
m2=9595.10

@@

<<echo=FALSE, results=tex>>=
# Default values:
t1=10449.670
t2=NA
m1=1261.6
m2=NA

if(Sys.getenv("RUN_EXAMPLES")=="TRUE")
{
  # simulated data, replicate rows 27x
}
require(xtable)
ans = data.frame(Method=c("tcor","brute force"), `Wall-clock time (s)`=c(t1, t2), `Peak excess memory use (MB)`=c(m1, m2), check.names=FALSE)
print(
    xtable(ans,
        caption = "TCGA ACC Methylation450 (80 rows x 394,014 columns, threshold=0.99)", label = "TCGA3", caption.placement = "top")
  ,size="small", table.placement="ht", include.rownames=FALSE)
@

>>>>

The number of correlation coefficients grows quadratically with the number of
vectors. Simply computing all pairs of correlation coefficients and then
filtering out coefficients below a given threshold may not be computationally
feasible for high-dimensional data in modern genomics and other applications.
Considerable attention has been devoted to \emph{pruning
methods} that cheaply prune away all but the most highly-correlated pairs of
vectors.  For example the TAPER and related algorithms of Xiong et al. develop
pruning rules based on the frequency of mutually co-occurring observations (rows)
among vector pairs; see \cite{prune1, prune2} and the references therein.
Related methods approximate a set containing the most highly-correlated
pairs using hashing algorithms\cite{prune3}. A number of other methods arrive
at approximate sets of the most highly-correlated pairs using
dimensionality-reduction techniques like the discrete Fourier
transform\cite{timeseries}. The method of Wu, et al.\cite{svd-similarity},
conceptually very similar to our approach, uses the singular value
decomposition (SVD) to find all pairs of close vectors with respect to a
distance metric (instead of correlation).  Our method finds all of the most
highly-correlated vector pairs with respect to a given threshold by pruning
along an intuitive path of decreasing variance defined by the data and computed
by a truncated SVD.

@@

We evaluated the algorithm using three public gene expression and methylation
datasets.  Most tests were performed on a desktop PC with a single AMD
A10-7850K 3.7$\,$GHz Athlon quad-core CPU and 16$\,$GB 1333$\,$MHz unbuffered
RAM running the Ubuntu 12.04.5 GNU/Linux OS. Tests were performed with 64-bit R
version 3.1.2 using double-precision arithmetic and the OpenBLAS library
version 0.1alpha2.2-3 (based on Goto's BLAS version 1.13) with {\tt
OMP\_NUM\_THREADS=4}. The last experiment includes a test run on a Linux
cluster as described in the text.

All examples use the reference R implementation, {\tt tcor()}, which can be
installed for the R language from the development GitHub repository using the
{\tt devtools} package with:

@@

Anyone can quickly run the example to experiment with the algorithm and
compare its output with brute force results. The example uses the {\tt
BicatYeast} data from the {\tt biclust} R package and related {\tt BicAT}
software\cite{biclust, bicat}. The data consist of 70 experiments involving the
Saccharomyces Cerevisiae yeast measuring gene expression data across 419
Affymetrix probes, represented as a $70\times 419$ real-valued matrix.  The
example finds all pairs of columns (Affymetrix probes) that meet or exceed a
Pearson correlation value of 0.95.  We use a projected dimension $p=10$ in {\tt tcor()}.
\begin{small}
<<eval=TRUE,echo=TRUE>>=
library(irlba)
source("tcor.R")  # XXX Replace later with tcor pacakge
library(biclust)  # for the BicatYeast data
data(BicatYeast)  # in gene by sample orientation
A = t(BicatYeast) # to correlate genes to genes
threshold = 0.95

@@

<<echo=FALSE, results=tex>>=
# --------------------------------------------------------------------------------------------------
# The first time you run this, you need to download the data as outlined in the
# following commented-out section. This code creates a file named
# 'rpkm.RData' used subsequently.
#
# To run any of the remaining (large) examples, first set the system environment
# variable RUN_EXAMPLES=TRUE (otherwise they won't run and representative results
# from Bryan's test machine will be printed).
#
#library(doMC)
# Download and extract this file with, for example:
# wget  http://gdac.broadinstitute.org/runs/stddata__2015_08_21/data/BRCA/20150821/gdac.broadinstitute.org_BRCA.Merge_rnaseq__illuminahiseq_rnaseq__unc_edu__Level_3__gene_expression__data.Level_3.2015082100.0.0.tar.gz
# tar zxf gdac.broadinstitute.org_BRCA.Merge_rnaseq__illuminahiseq_rnaseq__unc_edu__Level_3__gene_expression__data.Level_3.2015082100.0.0.tar.gz
# cd gdac.broadinstitute.org_BRCA.Merge_rnaseq__illuminahiseq_rnaseq__unc_edu__Level_3__gene_expression__data.Level_3.2015082100.0.0
# R --slave < read.R   # (run this file)
#f = "BRCA.rnaseq__illuminahiseq_rnaseq__unc_edu__Level_3__gene_expression__data.data.txt"
#N = 20534  # total number of lines
#h = 2      # header lines
# Read the first header line with sample IDs:
#id = as.vector(read.table(f, sep="\t", stringsAsFactors=FALSE, header=FALSE, nrows=1))
# Read the 2nd header line with sample type (raw, RPKM, etc.)
#header = as.vector(read.table(f, sep="\t", stringsAsFactors=FALSE, skip=1, header=FALSE, nrows=1))
# Read the rest of the file in parallel. I prefer doMC/foreach over mclapply
# for this because foreach can combine results incrementally.
#cores = 4
#registerDoMC(cores)
#block = floor((N - h)/cores)
#t1 = proc.time()
#ans = foreach(j=1:cores, .combine=rbind) %dopar%
#{
#    skip = block * (j - 1) + h
#    nrows = ifelse(j == cores, -1, block)
#    read.table(f, sep="\t", stringsAsFactors=FALSE, skip=skip, header=FALSE, nrows=nrows)
#}
#names(ans) = header
#print(proc.time() - t1)
# (All timings shown for a quad-core Athlon home PC reading from an older
# OCZ REVODRIVE3 X2 SSD drive (ca 2014).)
#   user  system elapsed
#494.549   4.134 131.232
# Compare with sequential approach:
#t1 = proc.time()
#x = read.table(f, sep="\t", stringsAsFactors=FALSE, skip=1, header=TRUE)
#print(proc.time() - t1)
#   user  system elapsed
#517.430   1.318 519.553
# Create a samples x gene matrix of RPKM values
#genes = ans[, 1] # gene labels
#idx   = grep("RPKM", names(ans))    # Select columns with RPKM values
#rpkm     = t(as.matrix(ans[, idx]))
# Label the coordinates by sample ID and gene:
#rownames(rpkm) = id[idx]
#colnames(rpkm) = genes
#save(rpkm, file="rpkm.RData")
# --------------------------------------------------------------------------------------------------
# Default values:
t1=40.34
t2=35.60
m1=39.10
m2=9595.10

@@

We illustrate that step 4 of Algorithm \ref{prune} can be computed in parallel
by comparing the above results run on our reference test PC with results run on
a modest GNU/Linux cluster at Paradigm4, Inc. The cluster consisted of four
computers connected by 10 gigabit ethernet. Each computer contained two Intel
Xeon E5-2650 2$\,$GHz CPUs (16 physical CPU cores per computer) and 64$\,$GB
ECC registered RAM running at 1,600$\,$MHz running CentOS 6.6 and 64-bit R
version 3.2.2 with the reference R BLAS library.

No code change was required to run in parallel on the cluster. The reference R
{\tt tcor()} implementation uses R's {\tt foreach}\cite{foreach} framework to
run on arbitrary parallel configurations (``back ends''). We simply registered
a {\tt doRedis}\cite{doredis} parallel back end with 16 R workers per computer
prior to invoking {\tt tcor()}.

The results computed on the Linux cluster are marked ``tcor (64 core cluster)''
below.

<<echo=FALSE, results=tex>>=
# Default values:
t1=10449.670
t2=NA
t3=NA
m1=1261.6
m2=NA
m3=NA

if(Sys.getenv("RUN_EXAMPLES")=="TRUE")
{
  # simulated data, replicate rows 27x
}
require(xtable)
ans = data.frame(Method=c("tcor (4 core PC)","brute force","tcor (64 core cluster)"), `Wall-clock time (s)`=c(t1, t2, t3), `Peak excess memory use (MB)`=c(m1, m2, m3), check.names=FALSE)
print(
    xtable(ans,
        caption = "TCGA ACC Methylation450 (80 rows x 394,014 columns, threshold=0.99)", label = "TCGA3", caption.placement = "top")
  ,size="small", table.placement="ht", include.rownames=FALSE, NA.string="NA")
@
